Everything you wanted to know about PanDA but were truly truly afraid to ask

### 2017-03-06 PandaEvolution.tex
Second iteration IV

The next LHC run for taking data (Run 3) will require more resources than the
Worldwide LHC Computing Grid (WLCG) can possibly provide. Currently, PanDA WMS
uses more than 100,000 cores at over 100 Grid sites, with a peak performance of
0.3 petaFLOPS. This capacity will be sufficient for the planned analysis and
data processing, but it will be insufficient for the Monte Carlo (MC) production
workflow and any extra activity. To alleviate these challenges, ATLAS is engaged
in a program to expand the current computing model to include additional
resources such as the opportunistic use of supercomputers as well as commercial
and academic clouds.

@@

Modern supercomputers have been designed to mainly support parallel computation
that requires runtime communication. Job execution is parallelized across
multiple cores, each core calculating a small part of the problem and
communicating with other cores via a message passing interface (MPI).
Accordingly, supercomputers have large number of worker nodes, connected through
a high-speed, low-latency dedicated network. Each worker node has multicore
CPUs, usually augmented with massively parallel Graphics Processing Units (GPUs)
or other types of specialized coprocessors.

PanDA WMS has been designed to support distributed computing. Executing
Workloads or workflows involves concurrent and/or sequential runs of possibly
large amount of jobs, each requiring no or minimal parallelization and no
communication at runtime. Accordingly, computing infrastructure like WLCG have
been designed to aggregate large amount of computing resources across sites,
often dislocated across geographical locations. While each site may deploy
runtime message-passing capabilities, usually these are not used to perform
distributed computations.

There are at least two approaches to enable PanDA WMS to execute ATLAS jobs on
supercomputers: (i) using the subset of resources and capabilities shared by
both supercomputers and Grid; (ii) reconciling the parallel and distributed
computing paradigms by means of dedicated abstractions. The former is a
pragmatic approach that enables the execution of specific workloads by
prototyping single-point solutions. The latter approach is principled and better
suited for a production-grade solution, capable of supporting general-purpose
workloads and workflows on both supercomputers and Grid infrastructures. These
two approaches are not mutually exclusive: developing a single-point solution
gives the opportunity to better understand the problem space, supporting the
creation of abstractions for the integration of computing paradigms.

This section illustrates the design, architecture and execution process of a
prototype developed by the PanDA team to support execution of the Geant4 stage
of the ATLAS production Monte Carlo workflow on the Titan supercomputer, managed
by the Oak Ridge Leadership Computing Facility (OLCF) at the Oak Ridge National
Laboratory (ORNL). After an analysis of the results obtained and the lessons
learned, the following section introduces the design and first experimental
characterization of a next generation executor (NGE). NGE is designed to
abstract resources and capabilities, enabling the concurrent execution of both
parallel and distributed computing.

@@

The Titan supercomputer, current number two
% (number one from November 2012 until June 2013)
on the Top 500 list~\cite{top500}, is a Cray XK7 system with 18,688 compute
nodes and a total of 299,008 CPU cores. Each compute node has an AMD Opteron
CPU, a Nvidia Tesla GPU, 32 GB of RAM and no local storage, though a 16 GB RAM
disk can be set up. Compute nodes use Cray’s Gemini interconnect for inter-node
MPI messaging but have no network connection to the outside world. Titan is
served by the Spider II~\cite{spider2}, a Lustre filesystem with 32 PB of disk
storage, and by a 29 PB HPSS tape storage system. Titan’s worker nodes run
Compute Node Linux, a run time environment based on SUSE Linux Enterprise
Server.

@@

Titan offers to its users login nodes and data transfer nodes (DTNs). Users have
to log into one of these nodes via ssh to submit jobs to Titan's PBS scheduler.
Titan's authentication and authorization model is based on two-factor
authentication with a RSA SecurID key, generated every 30 seconds. Login nodes
and DTNs are the only nodes with out/inbound wide area network connectivity.
Work nodes have only local network access. Data staging is supported via the
DTNs, from the wide area network and the OLCF high performance storage system
(HPSS). DTNs are shared across all Titan's users that can schedule jobs to a
dedicates PBS batch system to automate data staging. Fair-share policies are in
place both for the batch system and for the use of DTN resources.

@@

Currently, no HEP application can benefit from Titan's GPUs but some
computationally-intensive and non memory-intensive tasks can be offloaded from
the Grid to Titan's large amount of cores. Further, when HEP tasks can be
partitioned to independent jobs, Titan work nodes can be used to execute up to
16 concurrent jobs' payload, one for each core. Given these constraints and
challenges, the type of task most suitable for execution on Titan is HEP event
generation. Event generations are mostly computational and stand-alone code,
requiring less than 1GB of RAM at runtime and with small input data
requirements. The event generation in ATLAS accounts for 10--15\% of all jobs on
the Grid, corresponding to 2/3 of the current Grid capacity. A more difficult
workload to adapt for Titan would be that for detector simulation. Simulation
tasks tend to be tied to the framework of a particular simulation, often
requiring database access for geometry and relevant run conditions.

\mtnote{Currently, is PanDA running event generation, simulation or both on Titain?}
\mtnote{Please provide a diagram illustrating event generation and simulation
stats compared to the whole ATLAS computation activity.}

Event generation and detector simulation are part of the ATLAS production Monte
Carlo (MC) workflow (also known as MC production chain)~\cite{mcworkflow}. The
MC workflow consists of four main stages: event generation, detector simulation,
digitization, and reconstruction. Even generation creates sets of particle
four-momenta via different generators, e.g., PYTHIA~\cite{herwig} and
HERWIG~\cite{herwig}. The detector simulator is called Geant4~\cite{geant4} and
simulates the interaction of these particles with the sensitive material of the
ATLAS detector. Each interaction creates a so-called hit and all hits are
collected and passed to the digitalization stage. There, hits are further
process to mimic the readout of the detector, including subtle events specific
to each detector's sub-component. Finally, the reconstruction stage operates
local pattern recognition, creating high-level objects like particles and jets.

@@

The unavailability of pilots is a potential drawback when executing distributed
workloads like MC event generation and simulation. Pilots are used to increase
the throughput of distributed workloads: while pilots have to wait in the
supercomputer's queue, once scheduled, they can pull and execute jobs avoiding
the supercomputer's queue. This is particularly relevant for machines like Titan
where queue policies privilege parallel jobs on the base of the number of work
nodes they request: the higher the number of nodes, the shorter the amount of
queue time (modulo fair share policies).

@@

Titan's backfill functionality offers the opportunity not to use the pilot
abstraction while still avoiding queue time overheads. At any point in time,
Titan's Moab scheduler shows how many work nodes cannot be used by the jobs
already queued: All these jobs are either too large or their walltime is too
long. By submitting a job requesting the number of free work nodes and walltime,
queue time can be avoided.

Compared to pilots, backfill has the disadvantage of limiting the amount of work
nodes that can be requested. Pilots are normal jobs: they can request as many
work nodes for as much time as a queue can offer. On the contrary, jobs sized on
the base of the backfill availability depend on the amount of work nodes that
cannot be given to any other job. Usually, this amount of work nodes is a
fraction of the total capacity of the queue but the size of Titan mitigates this
limitation. Every year, about 10\% of Titan's capacity remains
unused~\cite{titan_utilization}, for a total of 30,000 cores that correspond to
270M core hours per year, roughly 30\% of the overall capacity of WLCG.

@@

Accordingly, PanDA Pilot has been redesigned to serve as a job broker On Titan.
Architected as a stand-alone service to be executed on a DTN, this PanDA Broker
implements functionalities to: (i) interrogate Titan about backfill
availability; (ii) pull MC jobs and events; (iii) wrap these jobs' payload into
PBS MPI jobs; (iv) submitting PBS jobs to Titan's batch system; and (v) staging
input/output files. Backfill querying, payload wrapping, and job submission
required a new implementation while job and events pulling, and file staging
were mostly inherited from PanDA Pilot.

@@

MPI wrappers are submitted to Titan's PBS batch system via
RADICAL-SAGA~\cite{radical-saga}. The Simple API for Grid Applications (SAGA)
offers a unified interface to diverse job schedulers and file transferring
services. In this way, SAGA provides an interoperability layer that lowers the
complexity of using distributed infrastructure. RADICAL-SAGA is a Python module,
compliant with the OGF GFD.90 SAGA specification~\cite{saga-spec}. Behind the
API façade, RADICAL-SAGA implements a adaptor architecture: each adaptor
interface the SAGA API with different middleware systems and services, including
the batch scheduler of Titan.

The data staging capabilities of the PanDA Broker are implemented via a file
system that is shared among DTNs and work nodes. The input files required by the
payload are downloaded via the ATLAS DDM service and then stored in a directory
on the shared filesystem. The MPI wrapper setup process includes making the
location of these files known to the payload. The PanDA Broker can locate the
Payloads' output files on the shared filesystem and transfer them from Titan to
the ATLAS Tier 1 computing center at Brookhaven National Lab (BNL) or any other
Grid site.

Once deployed, every PanDA Broker supports the execution of MC event generation
and detector simulation on Titan in 10 steps: (i) PanDA Broker queries Titan's
Moab scheduler about currently unused worker nodes;
% and checks whether the time and size of the available backfill resources are
% suitable to execute jobs;
(ii) The PanDA Broker transmits this information to the Job Dispatcher module of
the PanDA server and, in response, it gets a list of jobs that have been bound
to Titan by JEDI; (iii) based on the jobs attributes, the PanDA Broker requests
to the Data Service module of the PanDA Server to transfer the necessary input
data to the DTN via the ATLAS DDM service; (iv) once all the data are
transferred, the PanDA Broker calculates the number of cores and duration of
jobs that can be submitted to Titan on the base of the information collected
about backfill worker nodes availability; (v) the PanDA Broker creates one or
more PBS MPI scripts to wrap enough jobs' payload to fit the work nodes
available via backfill; (vi) the PanDA Broker submits these MPI jobs to the
Titan's PBS batch system via RADICAL-SAGA; (vii) the MPI jobs waits a minimal
amount of time into the queue until are scheduled for execution; (viii) upon
execution on the worker nodes, each MPI script sets up the execution environment
required by the wrapped payload, assembling a specific AthenaMP configuration in
ramdisk; (ix) the payload is executed into dedicated processes, 16 for each work
node, 1 for each core; (x) the PanDA Broker locates the output files produced by
the payload on the shared filesystem and transfer them to BNL ATLAS Tier 1
computing centre.

>>>>

The next LHC run for taking data (Run 3) will require more resources than the
Worldwide LHC Computing Grid (WLCG) can possibly provide. Currently, PanDA WMS
uses more than 100,000 cores at over 100 Grid sites, with a peak performance of
0.3 petaFLOPS. This capacity will be sufficient for the planned analysis and
data processing, but it will be insufficient for the Monte Carlo production
workflow and any extra activity. To alleviate these challenges, ATLAS is engaged
in a program to expand the current computing model to include additional
resources such as the opportunistic use of supercomputers as well as commercial
and academic clouds.

@@

Modern supercomputers have been designed mainly to support parallel computation
that requires runtime communication. Job execution is parallelized across
multiple cores, each core calculating a small part of the problem and
communicating with other cores via a message passing interface (MPI).
Accordingly, supercomputers have large number of worker nodes, connected through
a high-speed, low-latency dedicated network. Each worker node has multicore
CPUs, usually augmented with massively parallel Graphics Processing Units (GPUs)
or other types of specialized coprocessors.

PanDA WMS has been designed to support distributed computing. Executing ATLAS
workloads or workflows involves concurrent and/or sequential runs of possibly
large amount of jobs, each requiring no or minimal parallelization and no
runtime communication. Thus, computing infrastructure like WLCG have been
designed to aggregate large amount of computing resources across sites, often
geographically dislocated. While each site may deploy runtime message-passing
capabilities, usually these are not used to perform distributed computations.

There are at least two approaches to enable PanDA WMS to execute ATLAS workloads
or workflows on supercomputers: (i) using the subset of resources and
capabilities shared by both supercomputers and WLCG; (ii) reconciling the
parallel and distributed computing paradigms by means of dedicated abstractions.
The former is a pragmatic approach that enables the execution of specific
workloads by prototyping single-point solutions. The latter is a principled
approach, better suited for a production-grade solution, capable of supporting
general-purpose workloads and workflows on both supercomputers and Grid
infrastructures. These two approaches are not mutually exclusive: developing a
single-point solution gives the opportunity to better understand the problem
space, supporting the creation of abstractions for the integration of computing
paradigms.

This section illustrates the design and architecture of a job broker prototyped
by the PanDA team. This broker supports execution of part of the ATLAS
production Monte Carlo workflow on Titan, a leadership-class supercomputer
managed by the Oak Ridge Leadership Computing Facility (OLCF) at the Oak Ridge
National Laboratory (ORNL). After an analysis of the results obtained and the
lessons learned, the following section introduces the design and first
experimental characterization of a next generation executor (NGE). NGE is
designed to abstract resources and capabilities, enabling the concurrent
execution of both parallel and distributed computing on generic high performance
computing (HPC) machines.

@@

The Titan supercomputer, current number three
% (number one from November 2012 until June 2013)
on the Top 500 list~\cite{top500}, is a Cray XK7 system with 18,688 work nodes
and a total of 299,008 CPU cores. Each work node has an AMD Opteron CPU, a
Nvidia Tesla GPU, 32 GB of RAM and no local storage, though a 16 GB RAM disk can
be set up. Work nodes use Cray’s Gemini interconnect for inter-node MPI
messaging. Titan is served by the Spider II~\cite{spider2}, a Lustre filesystem
with 32 PB of disk storage, and by a 29 PB HPSS tape storage system. Titan’s
work nodes run Compute Node Linux, a run time environment based on SUSE Linux
Enterprise Server.

@@

Titan offers to its users login nodes and data transfer nodes (DTNs). Users have
to log into one of these nodes via ssh to submit jobs to Titan's PBS scheduler.
Titan's authentication and authorization model is based on two-factor
authentication with a RSA SecurID key, generated every 30 seconds. Login nodes
and DTNs are the only nodes with out/inbound wide area network connectivity.
Work nodes have only local network access. Data staging is supported via the
DTNs, from the wide area network and the OLCF high performance storage system
(HPSS). DTNs are shared across all Titan's users that can schedule jobs to a
dedicates batch system to automate data staging. Fair-share policies are in
place both for the batch system and for the use of DTN resources.

@@

Currently, no HEP application can benefit from Titan's GPUs but some
computationally-intensive and non memory-intensive tasks can be offloaded from
the Grid to Titan's large amount of cores. Further, when HEP tasks can be
partitioned to independent jobs, Titan work nodes can be used to execute up to
16 concurrent jobs' payload, one for each available core. Given these
constraints and challenges, the types of task most suitable for execution on
Titan is Monte Carlo event generation and detector simulation. Both types of
task are mostly computational and stand-alone code, requiring less than 1GB of
RAM at runtime and with small input data requirements. The event generation in
ATLAS accounts for 10--15\% of all jobs on WLCG, corresponding to 2/3 of the
current Grid capacity. Detector simulation accounts for \ldots .

\mtnote{Currently, is PanDA running event generation, detector simulation or
both on Titain?}
\mtnote{Please provide a diagram illustrating event generation and detector
simulation stats compared to the whole ATLAS computation activity.}

Event generation and detector simulation are part of the ATLAS production Monte
Carlo (MC) workflow (also known as MC production chain)~\cite{mcworkflow}. The
MC workflow consists of four main stages: event generation, detector simulation,
digitization, and reconstruction. Even generation creates sets of particle
four-momenta via different generators, e.g., PYTHIA~\cite{herwig} and
HERWIG~\cite{herwig}. The detector simulator is called Geant4~\cite{geant4} and
simulates the interaction of these particles with the sensitive material of the
ATLAS detector. Each interaction creates a so-called hit and all hits are
collected and passed on for digitalization where hits are further process to
mimic the readout of the detector. Finally, reconstruction operates local
pattern recognition, creating high-level objects like particles and jets.

@@

The unavailability of pilots is a potential drawback when executing distributed
workloads like MC event generation and detector simulation. Pilots are used to
increase the throughput of distributed workloads: while pilots have to wait in
the supercomputer's queue, once scheduled, they can pull and execute jobs
independently from the system's queue. This is particularly relevant for
machines like Titan where queue policies privilege parallel jobs on the base of
the number of work nodes they request: the higher the number of nodes, the
shorter the amount of queue time (modulo fair share policies).

@@

Titan's backfill functionality offers the opportunity not to use the pilot
abstraction while still avoiding queue time overheads. Backfill availability is
the number of work nodes that cannot be used by any of the jobs already queued
on Titan: All queued jobs are either too large or their walltime is too long. At
any point in time, Titan's Moab scheduler can be queried for backfill
availability and a job submitted shortly after placing this query and requesting
for no more than the backfill availability, is usually scheduled immediately
spending almost no time in the queue.

Compared to pilots, backfill has the disadvantage of limiting the amount of work
nodes that can be requested. Pilots are normal jobs: they can request as many
work nodes for as much time as a queue can offer. On the contrary, jobs sized on
the base of the backfill availability depend on the amount of work nodes that
cannot be given to any other job. Usually, backfill availability is a fraction
of the total capacity of the queue but the size of Titan mitigates this
limitation. Every year, about 10\% of Titan's capacity remains
unused~\cite{titan_utilization}, for a total of 30,000 cores, corresponding to
around 270M core hours per year, roughly 30\% of the overall capacity of WLCG.

@@

Given the communication requirements of PanDA Pilots and the unused capacity of
Titan, PanDA pilot was redesigned to serve as a job broker on Titan. Architected
as a stand-alone service to be executed on a DTN, this prototype called `PanDA
Broker' implements functionalities to: (i) interrogate Titan about backfill
availability; (ii) pull MC jobs and events; (iii) wrap jobs' payload into PBS
MPI jobs; (iv) submitting MPI PBS jobs to Titan's PBS batch system; and (v)
staging input/output files. Backfill querying, payload wrapping, and job
submission required a new implementation while job and events pulling, and file
staging were mostly inherited from PanDA Pilot.

@@

MPI wrappers are submitted to Titan's PBS batch system via
RADICAL-SAGA~\cite{radical-saga}, a Python module, compliant with the OGF GFD.90
SAGA specification~\cite{saga-spec}. The Simple API for Grid Applications (SAGA)
offers a unified interface to diverse job schedulers and file transferring
services. In this way, SAGA provides an interoperability layer that lowers the
complexity of using distributed infrastructures. Behind the API façade,
RADICAL-SAGA implements a adaptor architecture: each adaptor interface the SAGA
API with different middleware systems and services, including the batch
scheduler of Titan.

The data staging capabilities of the PanDA Broker are implemented via a file
system that is shared among DTNs and work nodes. The input files required by the
payload are downloaded via the ATLAS DDM service and then stored in a directory
on the shared filesystem. The MPI wrapper setup process includes making the
location of these files available to the payload. The PanDA Broker can locate
the Payloads' output files on the shared filesystem and transfer them from Titan
to the ATLAS Tier 1 computing center at Brookhaven National Lab (BNL) or any
other Grid site.

Once deployed, every PanDA Broker supports the execution of MC event generation
and detector simulation on Titan in 9 steps: (i) PanDA Broker queries Titan's
Moab scheduler about currently unused worker nodes;
% and checks whether the time and size of the available backfill resources are
% suitable to execute jobs;
(ii) the PanDA Broker queries the Job Dispatcher module of the PanDA server for
jobs that have been bound to Titan by JEDI; (iii)  the PanDA Broker requests to
the Data Service module of the PanDA Server what input data these jobs require
and transfer them to the DTN via the ATLAS DDM service; (iv) the PanDA Broker
creates one or more PBS MPI script, wrapping enough jobs' payload to fit
backfill availability; (v) the PanDA Broker submits these MPI jobs to the
Titan's PBS batch system via RADICAL-SAGA; (vi) the MPI jobs waits a minimal
amount of time into the queue until they are scheduled for execution; (vii) upon
execution on the worker nodes, each MPI script creates a ramdisk and sets up the
specific AthenaMP configuration required by the wrapped payload; (viii) the
payload is executed into dedicated processes, 16 for each work node, 1 for each
available core; (ix) upon completion of an MPI job, the PanDA Broker locates the
produced output files on the shared filesystem and transfer them to BNL ATLAS
Tier 1 computing centre.
